---
title: "Specifying Data Columns"
author: "Ryan M. Moore"
date: "2019-09-13"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using-biplotr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Import needed libraries.

```{r setup}
library(ggplot2)
library(biplotr)
```

# Specifying data columns

If you pass a `data.frame` to `pca_biplot` that has additional variables (e.g., factors with covariate info) that are not meant to be used in the PCA calculation, you must be sure to specify the `data_cols` parameter properly.

Let's use a modified version of the `iris` dataset to see what I mean.

```{r}
iris2 <- cbind(
  iris, 
  Group = c(
    rep(1, times = nrow(iris) / 2),
    rep(2, times = nrow(iris) / 2)
  )
)

print(str(iris2))
```

This data, `iris2` contains an additional column, `Group`, that groups the data into two groups, `1` and `2`.  However, as you can see, it is a `numeric` column.

```{r}
class(iris2$Group)
```

If we are not careful, the `Group` variable will be included in the PCA calculation even though for our data, it is a covariate that isn't meant to be included.

```{r}
p <- biplotr::pca_biplot(data = iris2, 
                         point_color = "Species", 
                         arrow_labels = TRUE,
                         limits_nudge_x = 1,
                         limits_nudge_y = -1)
plot(p$biplot)
```

As you can see, there is an extra variable in the biplot labeled `Group`.  This is because by default, `pca_biplot` treats every column other than the one specified by `point_color` as input to the PCA.  Since there are more covariates than just the one we specify in `point_color`, we need to manually specify the `data_cols` parameter like this:

```{r}
p <- biplotr::pca_biplot(data = iris2, 
                         data_cols = 1:4,
                         point_color = "Species", 
                         arrow_labels = TRUE,
                         limits_nudge_x = 1,
                         limits_nudge_y = -1)
plot(p$biplot)
```

What happens if we don't specify `data_cols` but had specified `Group` as a factor instead of numeric?  

```{r eval=FALSE, include=TRUE}
iris2$Group <- as.factor(iris2$Group)

p <- biplotr::pca_biplot(data = iris2, 
                         point_color = "Species", 
                         arrow_labels = TRUE,
                         limits_nudge_x = 1,
                         limits_nudge_y = -1)
```

In this case, an error occurs with a message somthing like this:

```
 Error in biplotr::pca_biplot(data = iris2, point_color = "Species", arrow_labels = TRUE, : All columns used for PCA are not numeric! Make sure to select only numeric columns for PCA! (cols included in PCA were: '1, 2, 3, 4, 6')
```

alerting you that the PCA was attempted on columns 1-4 and 6, which is not what we wanted.
